---
title: "Analysis on BRFSS2013"
author: "Guolong Mo"
date: "04/01/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 12)
load('/Users/babemomo/Dropbox/Coursera/Introduction to Probability and Data with R/CourseraIntroR/brfss2013.Rdata')
```

# Introduction
This project aims to analyze the Behavioral Risk Factor Surveillance System (BRFSS) 2013 dataset through three concrete research questions (See research question part).  
According to the BRFSS data user guide(https://www.cdc.gov/brfss/data_documentation/pdf/UserguideJune2013.pdf), BRFSS is "a cross-sectional telephone survey that state health departments conduct monthly over landline and cellular telephones with standardized questionnaires". **In other words, BRFSS 2013 is an observatory study to investigate the association among variables rather than the causal link**. The sample method is also detailed in the foregoing website. In brief, for landline telephone interviewers which accounted for ~80% of the questionnaires, a stratified strategy was applied based on the geographic division at multiple levels (termed "Disproportionate stratified sampling" in the document)  At lowest geographic level (distinguished by phone area number), the interviewers were randomly selected from the population. For cellular interviewers which account for ~20% of the questionnaires, simple random sampling was conducted at the state's level. **In other words, the sampling of BRFSS 2013 is randomized and can be generalized to the population at large.**  
In order to evaluate the non-response bias, the BRFSS protocol introduces a disposition code to score each interview attempt. The data quality report which includes response rates is available online (https://www.cdc.gov/brfss/annual_data/2013/pdf/2013_dqr.pdf).   
As suggested by the data quality report, I quote the following claim:  

> The response rate is the number of respondents who completed the survey as a proportion of all eligible and likely-eligible persons. The median survey response rate for all states, territories and Washington, DC, in 2013 was 46.4, and ranged from 29.0 to 60.3.For detailed information see the BRFSS Summary Data Quality Report.  

Despite the low response rate, BRFSS applied a "raking" weight to scale the data based on demographic characteristics of respondents. The report claimed this ensured accurate estimates for most measures.  
My project will be focused on the national-wide interpretation of BRFSS data in the U.S., so only only the data of 50 states will be included (i.e. data from the District of Columbia, Puerto Rico, and Guam are excluded). The analysis will be also focused on the data of main survey modules, since optional modules are not investigated by every state. 

# Research questions  
## 1.Is vegetarian diet inversely correlated with diabetes condition?  
Diet is an well-recognized factor for diabetes. Vegetarian diet, which mainly consists of vegetable and fruit, is often recommended for diabetes patients. Hence, it is interesting to investigate if there is any association between vegetable/fruit consumptions and diabetes conditions using the BRFSS 2013 dataset.  

* Three variables are involved (the variable names used in table are indicated in parenthesis): Diabetes (*diabete3*), Total Vegetables Consumed Per Day (*X_vegesum*), Total Fruits Consumed Per Day (*X_frutsum*).  
* **I will first select the variables of interest and filter out the missing observation using *tidyverse* package; and then summarize the vegetable/fruit consumption using the *summary()* function; at last I will visualize the summary using box plot from the *ggplot2* package.**

### Summary statistics  
```{r, message = FALSE}
library(tidyverse)
df1 <- brfss2013 %>%
  select(diabete3, X_vegesum, X_frutsum) %>% # Select the variables of interest.
  drop_na() %>% # Remove missing observations.
  filter(diabete3 != 'Yes, but female told only during pregnancy', diabete3 != 'No, pre-diabetes or borderline diabetes') # Remove diabetes observed only at pregnancy or in borderline conditions.
```
Summary of vegetable consumption of those who in diabetes condition
```{r}
summary(df1$X_vegesum[df1$diabete3 == 'Yes'])
```
Summary of vegetable consumption of those who in non-diabetes condition
```{r}
summary(df1$X_vegesum[df1$diabete3 == 'No'])
```

Summary of fruit consumption of those who in diabetes condition
```{r}
summary(df1$X_frutsum[df1$diabete3 == 'Yes'])
```

Summary of fruit consumption of those who in non-diabetes condition
```{r}
summary(df1$X_frutsum[df1$diabete3 == 'No'])
```
In the comparison of vegetable consumptions, diabetes patients tend to consume less than those who are in non-diabetes conditions (mean: 174.8 vs. 192.6; median: 150.0 vs. 167.0).  
In the comparison of fruit consumptions, diabetes patients tend to consume less than those who are in non-diabetes conditions (mean: 130.0 vs. 141.3; median: 100.0 vs. 104.0).  
In other words, non-diabetes conditions are associated with higher consumptions of vegetable and fruit.  

It is worthy noting that there are some extreme values which are 100 times higher than the average; hence the median may be a better indicator. Box plots will be use to visualize the the vegetable/fruit consumptions against different diabetes conditions.  

### Box plot for data visualization  
```{r, warning = FALSE, message = FALSE}
library(ggplot2)
library(cowplot)
gg1 <- ggplot(df1, aes(x = diabete3, y = X_vegesum, colour=diabete3)) +
  geom_boxplot() +
  scale_y_log10() + 
  ggtitle("Diabetes condition vs. Vegetable consumption") + 
  labs(x = "Diabetes condition", y = "Vegetable consumption (mg/day, log scale)", colour = "Diabetes") + 
  theme_cowplot()
gg2 <- ggplot(df1, aes(x = diabete3, y = X_frutsum, colour=diabete3)) +
  geom_boxplot() +
  scale_y_log10() + 
  ggtitle("Diabetes condition vs. Fruit consumption") + 
  labs(x = "Diabetes condition", y = "Fruit consumption (mg/day, log scale)", colour = "Diabetes") + 
  theme_cowplot()
plot_grid(gg1, gg2)
```

